Collective excitations in competing phases in two and three dimensions

### 2024-01-12 paper/main.tex
Läuft

The Hubbard model has been studied in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered,
which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.

@@

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_SC_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The left edge of each plot is at $-0.05$ rather than $0$ in order to improve the visibility of the phase peak.
    The shaded area marks the two-particle continuum.
    For each plot, the parameters are set to $U=-2.5t$ and $N_\gamma = 3000$, while $V$ is varied according to the text in the individual panels.
    The panels show the spectral functions in the (a) pure SC phase, (b) SC/CDW-coexistence, and (c) and (d) pure CDW phase, respectively.}
    \label{fig:resolvent_overview_SC}
\end{figure*}

@@

To classify the occurring peaks that lie outside the two-particle continuum, we turn towards the real part of the Green's function.
We plot it in close proximity to the peaks with a double logarithmic scale and fit the result linearly $y = ax + b$ to obtain its power-law behavior.
The real part is related to the imaginary part via the Kramers-Kronig relations.
Specifically, if the imaginary part is a $\delta$ distribution, the real part will be proportional to $1/\omega$, i.e., $a=-1$, and the peak has the weight $w_0 = \exp(b)$.
Furthermore, an exponent of $a=-2$ corresponds to the derivative of the $\delta$ distribution, i.e., $\delta'(\omega)$.
The fits themselves can be found in appendix \autoref{sec:fit_greens_functions}.

@@

\textcolor{red}{Ich spekuliere mal...}
The singularity in $\spectral{Higgs}$ behaves as $1/\sqrt{\omega - 2 \Delta_\text{SC}}$.
Previous studies on the dynamics of the order parameter found damped oscillations after a quench.
These have the frequency $\Omega = 2 \Delta_\text{SC}$ and fall off as $1/\sqrt{t}$ \cite{Yuzbashyan06}.
An inverse Fourier transform of $\spectral{Higgs}$ yields precisely the same behavior, thereby reinforcing the validity of our results.

In the CDW phase, see panels (c) and (d), both of $\spectral{Higgs}$ and $\spectral{Phase}$ become identical and form a $\delta$ peak below the two-particle continuum.
Following the Landau theory for continuous phase transitions, this is to be expected as the free energy is expanded in terms of the order parameter as
\begin{equation}
    f(\Delta_\text{SC}) \approx r |\Delta_\text{SC}|^2 + \frac{u}{2} |\Delta_\text{SC}|^4\,,
\end{equation}
with $r > 0$ and $u > 0$ if the system is outside of the superconducting phase. \textcolor{red}{Zitieren!}
In this case, the only minimum is at $\Delta_\text{SC} = 0$. 
That means that both possible modes have to overcome the same energy to be excited at all. \textcolor{red}{Passt die Beschreibung so?}

@@

At $V=0$, both the SC and the CDW phase coexist. 
Since the degeneracy of both phases is exact due to the SO(4) symmetry \cite{yang90}, 
there should not be any difference between $\spectral{Higgs}$ and $\spectral{CDW}$ as both spectral functions capture the amplitude fluctuations of their respective order parameters.
This is precisely what can be seen in panel (b) of \autoref{fig:resolvent_overview_SC}.
Both spectral functions have a peak at $\omega = 0$ and a singularity at the lower edge of the two-particle continuum $\omega = 2\Delta_\text{tot}$, with $\Delta_\text{tot} \coloneqq \sqrt{|\Delta_\text{CDW}|^2 + |\Delta_\text{SC}|^2}$.

\textcolor{red}{Ich spekuliere mal...}
In this phase, the mean-field single-particle energies are given by
\begin{equation}
    E_{\pm} = \pm \sqrt{\epsilon^2 + |\Delta_\text{CDW}|^2 + |\Delta_\text{SC}|^2}\,.
\end{equation}
This means that the proper order parameter is $\Delta_\text{tot}$.
Here, $\Delta_\text{CDW}$ and $\Delta_\text{SC}$ can change arbitrarily without affecting the system's energy as long as $\Delta_\text{tot}$ remains constant.
Thus, we associate the peak at $\omega=0$ with this shifting of order parameters.
The singularity at the lower edge of the two-particle continuum can be interpreted in a similar manner as the Higgs mode in a pure superconductor.
It captures the change of $\Delta_\text{tot}$ by changing either one of the composing order parameters.

@@

In the SC phase, $\spectral{CDW}$ features a $\delta$ peak below the continuum corresponding to the finite energy necessary to excite this kind of electronic density modulation.
For moderate-sized values of $V>0$, i.e., in the CDW phase, see panel (c) at $V=0.1t$, the aforementioned spectral function has a singularity at $\omega=2\Delta_\text{CDW}$.
This mode becomes hard and moves outside the two-particle continuum as $V$ is increased, see panel (d) at $V=0.5t$.
\textcolor{red}{Ich muss mal suchen, ob sowas schonmal beobachtet wurde.}

@@

>>>>

\textcolor{red}{Intro auf die Phasen und Anregungen beziehen, danach Theorie auspacken}
The Hubbard model has been studied in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact SO(4) symmetry was discovered,
which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_SC_CDW.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The left edge of each plot is at $-0.05$ rather than $0$ in order to improve the visibility of the phase peak.
    The shaded area marks the two-particle continuum.
    For each plot, the parameters are set to $U=-2.5t$ and $N_\gamma = 3000$, while $V$ is varied according to the text in the individual panels.
    The panels show the spectral functions in the (a) SC phase and the (b and c) CDW phase, respectively.}
    \label{fig:resolvent_overview_SC}
\end{figure*}

@@

@@

To classify the occurring peaks that lie outside the two-particle continuum, we turn toward the real part of the Green's function.
We plot it in close proximity to the peaks with a double logarithmic scale and fit the result linearly $y = ax + b$ to obtain its power-law behavior.
The real part is related to the imaginary part via the Kramers-Kronig relations.
Specifically, if the imaginary part is a $\delta$ distribution, the real part will be proportional to $1/\omega$, i.e., $a=-1$, and the peak has the weight $w_0 = \exp(b)$.
Furthermore, an exponent of $a=-2$ corresponds to the derivative of the $\delta$ distribution, i.e., $\delta'(\omega)$.
The fits themselves can be found in appendix \autoref{sec:fit_greens_functions}.

@@

The singularity in $\spectral{Higgs}$ behaves as $1/\sqrt{\omega - 2 \Delta_\text{SC}}$.
Previous studies on the dynamics of the order parameter found damped oscillations after a quench.
These have the frequency $\Omega = 2 \Delta_\text{SC}$ and fall off as $1/\sqrt{t}$ \cite{Yuzbashyan06}.
An inverse Fourier transform of $\spectral{Higgs}$ yields precisely the same behavior, thereby reinforcing the validity of our results.

In the CDW phase, see panels (b) and (c), both of $\spectral{Higgs}$ and $\spectral{Phase}$ become identical and form a $\delta$ peak below the two-particle continuum.
Following the Landau theory for continuous phase transitions, this is to be expected as the free energy is expanded in terms of the order parameter as
\begin{equation}
    f(\Delta_\text{SC}) \approx r |\Delta_\text{SC}|^2 + \frac{u}{2} |\Delta_\text{SC}|^4\,,
\end{equation}
with $r > 0$ and $u > 0$ if the system is outside of the superconducting phase. \textcolor{red}{Zitieren!}
In this case, the only minimum is at $\Delta_\text{SC} = 0$. 
That means that both possible modes have to overcome the same energy to be excited at all. \textcolor{red}{Passt die Beschreibung so?}

In the SC phase, $\spectral{CDW}$ features a $\delta$ peak below the continuum corresponding to the finite energy necessary to excite this kind of electronic density modulation.
For moderate-sized values of $V>0$, i.e., in the CDW phase, see panel (b) at $V=0.1t$, the aforementioned spectral function has a singularity at $\omega=2\Delta_\text{CDW}$.
This mode becomes hard and moves outside the two-particle continuum as $V$ is increased, see panel (c) at $V=0.5t$.

@@

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/resolvent_overview_V0.pdf}
    \caption{Spectral functions with respect to the operators listed in \eqref{eqn:resolvent_bases}.
    The left column shows the results for the square lattice while the right column shows the results for the simple cubic lattice.
    The left edge of each plot is at $-0.05$ rather than $0$ in order to improve the visibility of the peaks at $\omega=0$.
    The shaded area marks the two-particle continuum.
    For each plot, the parameters are set to $U=-2.5t$, $V=0$, and $N_\gamma = 3000$.
    Since the ratio of $\Delta_\text{SC}$ to $\Delta_\text{CDW}$ may be chosen arbitrarily, we show in (a) the spectral functions for $\Delta_\text{CDW} = 0$ and in (b) for $\Delta_\text{SC} = 0$.}
    \label{fig:resolvent_overview_V0}
\end{figure*}

At $V=0$, both the SC and the CDW phase coexist. 
In this phase, the mean-field single-particle energies are given by
\begin{equation}
    E_{\pm} = \pm \sqrt{\epsilon^2 + |\Delta_\text{CDW}|^2 + |\Delta_\text{SC}|^2}\,.
\end{equation}
This means that the proper order parameter is $\Delta_\text{tot} \coloneqq \sqrt{|\Delta_\text{CDW}|^2 + |\Delta_\text{SC}|^2}$.
Here, $\Delta_\text{CDW}$ and $\Delta_\text{SC}$ can change arbitrarily without affecting the system's energy as long as $\Delta_\text{tot}$ remains constant.
Note that the degeneracy of both phases is exact due to the SO(4) symmetry \cite{yang90}.

In \autoref{fig:resolvent_overview_V0}, we show the spectral functions for $U=-2.5t$ and $V=0$.
The different panels correspond to different choices for $\Delta_\text{CDW}$ and $\Delta_\text{SC}$.
First, in panel (a), we set $\Delta_\text{CDW} = 0$.
This results in a purely superconducting phase with the same features as in \autoref{fig:resolvent_overview_SC} panel (a).
The peak in $\spectral{CDW}$ moves to a very small, yet finite, $\omega$.

Secondly, panel (b) shows the spectral functions for $\Delta_\text{SC} = 0$, i.e., in a purely charge-ordered phase.
Similar to before, there is not much change compared to \autoref{fig:resolvent_overview_SC} panel (b).
The spectral functions $\spectral{Higgs}$ and $\spectral{Phase}$ are still identical, however, both exhibit a sharp peak at $\omega=0$.
This peak is described by $\delta'(\omega)$.
\textcolor{red}{Warum hat Higgs nun einen Peak bei 0, aber CDW hatte einen bei endlichem $\omega$? Das müsste ja bedeuten, dass ein Cooper-Paar energtisch kostenfrei ist, aber ein CDW-Exziton kostet uns etwas Energie.}

Lastly, we chose $\Delta_\text{CDW} = \Delta_\text{SC}$ for panel (c).
Now, $\spectral{Higgs}$ and $\spectral{CDW}$ are identical.
Note, that this only occurs for this specific ratio of $\Delta_\text{CDW}$ and $\Delta_\text{SC}$, otherwise the occurring peaks differ in magnitude.
Both spectral functions have a peak at $\omega = 0$, described by $\delta'(\omega)$, and a singularity at the lower edge of the two-particle continuum $\omega = 2\Delta_\text{tot}$.
\textcolor{red}{TODO: Exponent}

We associate the peak at $\omega=0$ with this shifting of order parameters, i.e., diminishing one and increasing the other while keeping $\Delta_\text{tot}$ constant.
The singularity at the lower edge of the two-particle continuum can be interpreted similarly to the Higgs mode in a pure superconductor.
It captures the change of $\Delta_\text{tot}$ by changing either one of the composing order parameters.

@@

@@

Specifically, we learn that for $V \ll t$, the peak position follows $\omega_0 = \alpha \sqrt{V}$ and its weight follows $w_0 = \beta / \sqrt{V}$, where $\alpha$ and $\beta$ are some constants.
This allows us to write $w_0 = \beta / (\alpha \omega_0)$. Furthermore, we know that the spectral functions need to be perfectly antisymmetrical about $\omega = 0$, i.e. we can observe
\begin{align}
    \mathcal{A}_\text{Phase} (\omega \ll t) &= w_0 (\delta (\omega - \omega_0) - \delta (\omega + \omega_0)) \nonumber \\
        &= - \frac{2\beta}{\alpha} \frac{\delta (\omega + \omega_0) - \delta (\omega - \omega_0)}{2\omega_0} \,.
\end{align} 
Taking the limit of $V \to 0$ corresponds to the limit $\omega_0 \to 0$, i.e., leading to an expression reminiscent of symmetric derivative
\begin{equation}
    \lim_{h \to 0} \frac{f(x + h) - f(x - h)}{2h} = f'(x)\,,
\end{equation}
where the equality holds for any differentiable $f$.
Applying this to our case yields
\begin{align}
    \lim_{\omega_0 \to 0} \mathcal{A}_\text{Phase} (\omega \ll t) = - \frac{2 \beta}{\alpha} \delta'(\omega)\,.
\end{align}
This analysis reinforces our previous identification of the phase peak as $\delta' (\omega)$.
